### 2018-10-09 post-linearity-2018/sequent.tex
Added all proofs in appendix. Added equivalence between sequence calculus and natural deduction. Changed some notations for rule names.
Some small format changes.

We prove cut elimination for the sequent calculus. We define the
\textit{degree} $|X|$ (resp. $|A|$) of a commutative (resp. non-commutative)
formula to be the number of logical connectives in the proposition. For
instance, $|[[X (*) Y]]| = |[[X]]| + |[[Y]]| + 1$. The \textit{cut rank}
$c(\Pi)$ of a proof $\Pi$ is one more than the maximum of the ranks of all
the cut formulae in $\Pi$, and $0$ if $\Pi$ is cut-free. Then the
\textit{depth} $d(\Pi)$ of a proof $\Pi$ is the length of the longest path
in the proof tree (so the depth of an axiom is $0$). The key to the proof
of cut elimination is the following lemma, which shows how to transform a
single cut, either by removing it or by replacing it with one or more
simpler cuts.
\begin{lemma}[Cut Reduction]
  \label{lem:cut-reduction}
  \begin{enumerate}
  \item If $\Pi_1$ is a proof of $[[I |-c X]]$ and $\Pi_2$ is a proof of
  $[[P1,X,P2 |-c Y]]$ with $c(\Pi_1)$, $c(\Pi_2)\leq |X|$, then there exists
  a proof $\Pi$ of $[[P1, I, P2 |-c Y]]$ with $c(\Pi)\leq |X|$.
  \item If $\Pi_1$ is a proof of $[[I |-c X]]$ and $\Pi_2$ is a proof of
  $[[G1; X; G2 |-l A]]$ with $c(\Pi_1)$, $c(\Pi_2)\leq |X|$, then there
  exists a proof $\Pi$ of $[[G1; I; G2 |-l A]]$ with $c(\Pi)\leq |X|$.
  \item If $\Pi_1$ is a proof of $[[G |-l A]]$ and $\Pi_2$ is a proof of
  $[[D1; A; D2 |-l B]]$ with $c(\Pi_1)$, $c(\Pi_2)\leq |A|$, then there
  exists a proof $\Pi$ of $[[D1; G; D2 |-l B]]$ with $c(\Pi)\leq |A|$.
  \end{enumerate}
\end{lemma}
The proof for Lemma~\ref{lem:cut-reduction} is done case by case on the last
step of $\Pi_1$ and $\Pi_2$ and by induction on $d(\Pi_1)$ and $d(\Pi_2)$,
following \cite{Mellies:2009}. For instance, suppose $\Pi_1$ is a proof of
$[[I1, X2, X1, I2 |-c Y]]$ and $\Pi_2$ is a proof of $[[P1, Y, P2 |-c Z]]$. 
Consider the case where the last step in $\Pi_1$ uses the rule
$\NDdruleTXXbetaName{}$. $\Pi_1$ can be depicted as follows, where the
previous steps is denoted by $\pi$:
\begin{center}
  \scriptsize
  $\Pi_1$:
  \begin{math}
    $$\mprset{flushleft}
    \inferrule* [right={\tiny ex}] {
      {
        \begin{array}{c}
          \pi \\
          {[[I1, X1, X2, I2 |-c Y]]}
        \end{array}
      }
    }{[[I1, X2, X1, I2 |-c Y]]}
  \end{math}
\end{center}
By assumption, $c(\Pi_1),c(\Pi_2)\leq |Y|$. By induction on $\pi$ and
$\Pi_2$, there is a proof $\Pi'$ for the sequent \\
$[[P1, I1, X1, X2, I2, P2 |-c Z]]$ s.t. $c(\Pi')\leq|Y|$. Therefore, the
proof $\Pi$ can be constructed as follows, and $c(\Pi)=c(\Pi')\leq|Y|$.
\begin{center}
  \scriptsize
  \begin{math}
    $$\mprset{flushleft}
    \inferrule* [right={\tiny ex}] {
      {
        \begin{array}{c}
          \Pi' \\
          {[[P1, I1, X1, X2, I2, P2 |-c Z]]}
        \end{array}
      }
    }{[[P1, I1, X2, X1, I2, P2 |-c Z]]}
  \end{math}
\end{center}
The full proof of Lemma~\ref{lem:cut-reduction} is in the Appendix. Then we
have the following lemma.

\begin{lemma}
  \label{lem:less-cut-rank}
  Let $\Pi$ be a proof of a sequent $[[I |-c X]]$ or $[[G |-l A]]$ s.t.
  $c(\Pi)>0$. Then there is a proof $\Pi'$ of the same sequent with
  $c(\Pi')<c(\Pi)$.
\end{lemma}
\begin{proof}
  We prove the lemma by induction on $d(\Pi)$. We denote the proof $\Pi$ by 
  $\pi+r$, where $r$ is the last inference of $\Pi$ and $\pi$ denotes the
  rest of the proof. If $r$ is not a cut, then by induction hypothesis on
  $\pi$, there is a proof $\pi'$ s.t. $c(\pi')>c(\pi)$ and $\Pi'=\pi'+r$.
  Otherwise, we assume $r$ is a cut on a formula $X$. If $c(\Pi)>|X|+1$,
  then there is a cut on $|Y|$ in $\pi$ with $|Y|>|X|$. So we can apply
  the induction hypothesis on $\pi$ to get $\Pi'$ with $c(\Pi')<c(\Pi)$. The
  last case to consider is when $c(\Pi)=|X|+1$ (note that $c(\Pi)$ cannot be
  less than $|X|+1$). In this case, $\Pi$ is in the form of
  \begin{center}
    \scriptsize
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right={\tiny cut}] {
        {
          \begin{array}{cc}
            \Pi_1 & \Pi_2 \\
            {[[I |-c X]]} & {[[P1, X, P2 |-c Y]]}
          \end{array}
        }
      }{[[P1, I, P2 |-c Y]]}
    \end{math}
    \qquad\qquad
    or,
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right={\tiny cut1}] {
        {
          \begin{array}{cc}
            \Pi_1 & \Pi_2 \\
            {[[I |-c X]]} & {[[G1; X; G2 |-l A]]}
          \end{array}
        }
      }{[[G1; I; G2 |-l A]]}
    \end{math}
  \end{center}
  By assumption, $c(\Pi_1),c(\Pi_2)\leq |X|+1$. By induction, we can
  construct $c(\Pi_1')$ proving $[[I |-c X]]$ and $c(\Pi_2')$ proving
  $[[P1, X, P2 |-c Y]]$ (or $<[[G1; X; G2 |-l A]]$) with
  $c(\Pi_1'), c(\Pi_2')\leq |X|$. Then by Lemma~\ref{lem:cut-reduction}, we
  can construct $\Pi'$ proving $[[P1, I, P2 |-c Y]]$ (or
  $[[G1; I; G2 |-l A]]$) with $c(\Pi')\leq |X|$. \\
  The case where the last inference is a cut on a formula $A$ is similar as
  when it is a cut on $X$.
\end{proof}

By induction on $c(\Pi)$ and Lemma~\ref{lem:less-cut-rank}, the cut
elimination theorem follows immediately.
\begin{theorem}[Cut Elimination]
  Let $\Pi$ be a proof of a sequent $[[I |-c X]]$ or $[[G |-l A]]$ s.t.
  $c(\Pi)>0$. Then there is an algorithm which yields a cut-free proof
  $\Pi'$ of the same sequent.
\end{theorem}

>>>>

We prove cut elimination for the sequent calculus. We define the
\textit{degree} $|X|$ (resp. $|A|$) of a commutative (resp. non-commutative)
formula to be the number of logical connectives in the proposition. For
instance, $|[[X (*) Y]]| = |[[X]]| + |[[Y]]| + 1$. The \textit{cut rank}
$c(\Pi)$ of a proof $\Pi$ is one more than the maximum of the ranks of all
the cut formulae in $\Pi$, and $0$ if $\Pi$ is cut-free. Then the
\textit{depth} $d(\Pi)$ of a proof $\Pi$ is the length of the longest path
in the proof tree (so the depth of an axiom is $0$). The key to the proof
of cut elimination is the following lemma, which shows how to transform a
single cut, either by removing it or by replacing it with one or more
simpler cuts.
\begin{lemma}[Cut Reduction]
  \label{lem:cut-reduction}
  The cut-reduction steps are as follows:
  \begin{enumerate}
  \item If $\Pi_1$ is a proof of $[[I |-c X]]$ and $\Pi_2$ is a proof of
  $[[P1,X,P2 |-c Y]]$ with $c(\Pi_1)$, $c(\Pi_2)\leq |X|$, then there exists
  a proof $\Pi$ of $[[P1, I, P2 |-c Y]]$ with $c(\Pi)\leq |X|$.
  \item If $\Pi_1$ is a proof of $[[I |-c X]]$ and $\Pi_2$ is a proof of
  $[[G1; X; G2 |-l A]]$ with $c(\Pi_1)$, $c(\Pi_2)\leq |X|$, then there
  exists a proof $\Pi$ of $[[G1; I; G2 |-l A]]$ with $c(\Pi)\leq |X|$.
  \item If $\Pi_1$ is a proof of $[[G |-l A]]$ and $\Pi_2$ is a proof of
  $[[D1; A; D2 |-l B]]$ with $c(\Pi_1)$, $c(\Pi_2)\leq |A|$, then there
  exists a proof $\Pi$ of $[[D1; G; D2 |-l B]]$ with $c(\Pi)\leq |A|$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  This proof is done case by case on the last step of $\Pi_1$ and
  $\Pi_2$ and by induction on $d(\Pi_1)$ and $d(\Pi_2)$, following
  \cite{Mellies:2009}. For instance, suppose $\Pi_1$ is a proof of
  $[[I1, X2, X1, I2 |-c Y]]$ and $\Pi_2$ is a proof of $[[P1, Y, P2
      |-c Z]]$.  Consider the case where the last step in $\Pi_1$ uses
  the rule $\NDdruleTXXbetaName{}$. $\Pi_1$ can be depicted as
  follows, where the previous steps is denoted by $\pi$:
  \begin{center}
    \scriptsize
    $\Pi_1$:
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right={\tiny ex}] {
        {
          \begin{array}{c}
            \pi \\
                {[[I1, X1, X2, I2 |-c Y]]}
          \end{array}
        }
      }{[[I1, X2, X1, I2 |-c Y]]}
    \end{math}
  \end{center}
  By assumption, $c(\Pi_1),c(\Pi_2)\leq |Y|$. By induction on $\pi$ and
  $\Pi_2$, there is a proof $\Pi'$ for the sequent \\
  $[[P1, I1, X1, X2, I2, P2 |-c Z]]$ s.t. $c(\Pi')\leq|Y|$. Therefore, the
  proof $\Pi$ can be constructed as follows, and $c(\Pi)=c(\Pi')\leq|Y|$.
  \begin{center}
    \scriptsize
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right={\tiny ex}] {
        {
          \begin{array}{c}
            \Pi' \\
                 {[[P1, I1, X1, X2, I2, P2 |-c Z]]}
          \end{array}
        }
      }{[[P1, I1, X2, X1, I2, P2 |-c Z]]}
    \end{math}
  \end{center}
  The full proof of Lemma~\ref{lem:cut-reduction} is in
  Appendix~\ref{app:cut-reduction}. 
\end{proof}
\noindent
Then we have the following lemma.

\begin{lemma}
  \label{lem:less-cut-rank}
  Let $\Pi$ be a proof of a sequent $[[I |-c X]]$ or $[[G |-l A]]$ s.t.
  $c(\Pi)>0$. Then there is a proof $\Pi'$ of the same sequent with
  $c(\Pi')<c(\Pi)$.
\end{lemma}
\begin{proof}
  We prove the lemma by induction on $d(\Pi)$. We denote the proof $\Pi$ by 
  $\pi+r$, where $r$ is the last inference of $\Pi$ and $\pi$ denotes the
  rest of the proof. If $r$ is not a cut, then by induction hypothesis on
  $\pi$, there is a proof $\pi'$ s.t. $c(\pi')>c(\pi)$ and $\Pi'=\pi'+r$.
  Otherwise, we assume $r$ is a cut on a formula $X$. If $c(\Pi)>|X|+1$,
  then there is a cut on $|Y|$ in $\pi$ with $|Y|>|X|$. So we can apply
  the induction hypothesis on $\pi$ to get $\Pi'$ with $c(\Pi')<c(\Pi)$. The
  last case to consider is when $c(\Pi)=|X|+1$ (note that $c(\Pi)$ cannot be
  less than $|X|+1$). In this case, $\Pi$ is in the form of
  \begin{center}
    \scriptsize
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right={\tiny cut}] {
        {
          \begin{array}{cc}
            \Pi_1 & \Pi_2 \\
            {[[I |-c X]]} & {[[P1, X, P2 |-c Y]]}
          \end{array}
        }
      }{[[P1, I, P2 |-c Y]]}
    \end{math}
    \qquad\qquad
    or,
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right={\tiny cut1}] {
        {
          \begin{array}{cc}
            \Pi_1 & \Pi_2 \\
            {[[I |-c X]]} & {[[G1; X; G2 |-l A]]}
          \end{array}
        }
      }{[[G1; I; G2 |-l A]]}
    \end{math}
  \end{center}
  By assumption, $c(\Pi_1),c(\Pi_2)\leq |X|+1$. By induction, we can
  construct $c(\Pi_1')$ proving $[[I |-c X]]$ and $c(\Pi_2')$ proving
  $[[P1, X, P2 |-c Y]]$ (or $<[[G1; X; G2 |-l A]]$) with
  $c(\Pi_1'), c(\Pi_2')\leq |X|$. Then by Lemma~\ref{lem:cut-reduction}, we
  can construct $\Pi'$ proving $[[P1, I, P2 |-c Y]]$ (or
  $[[G1; I; G2 |-l A]]$) with $c(\Pi')\leq |X|$. \\
  The case where the last inference is a cut on a formula $A$ is similar as
  when it is a cut on $X$.
\end{proof}
\noindent
By induction on $c(\Pi)$ and Lemma~\ref{lem:less-cut-rank}, the cut
elimination theorem follows immediately.
\begin{theorem}[Cut Elimination]
  Let $\Pi$ be a proof of a sequent $[[I |-c X]]$ or $[[G |-l A]]$ s.t.
  $c(\Pi)>0$. Then there is an algorithm which yields a cut-free proof
  $\Pi'$ of the same sequent.
\end{theorem}
